---
title: 'Reproducible Research: Peer Assessment 1'
author: "Tochi Okeke"
date: "December 8, 2018"
output:
  html_document: null
  'html_document:': default
  keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
### Reproducible Research Report 1
#### Analyzing data from the Activity Monitoring Dataset

First, I will set my working directory and download the data using the data.table package
```{r}
setwd("~/R/RepData_PeerAssessment1-master")
```

```{r}
library(data.table)#for subsetting
library(dplyr) #for data cleaning and manipulation
library(ggplot2) #for visualization
library(knitr)# to create the html report
opts_chunk$set(fig.path = "./figure/")
data <- fread("activity.csv")
head(data)
```
Taking a quick look at the data with summary

```{r}
summary(data)
```
Checking the class of the 'date' column
```{r}
class(data$date)
```
Downloading the lubridate library to convert the date column into 'date' class
```{r}
library(lubridate)
data$date <- ymd(data$date)
print(class(data$date))
```


### What is the mean total number of steps taken per day?

#### Calculate the total number of steps per day.
Using the data.table package, I will aggregate the data by date and find the total steps per day.
```{r}

totalsteps <- data[,sum(steps),date]
head(totalsteps)

```
The mean is 
```{r}
mean(totalsteps$V1,na.rm=TRUE)
``` 
The median is
```{r} 
median(totalsteps$V1,na.rm=TRUE)
```

```{r}
#Creating the histogram and adding labels
 hist(totalsteps$V1,xlab="Total Steps per Day",main="Histogram of Total Steps per Day",col="lightblue")

#Adding a vertical line with to indicate the mean
abline(v=c(mean(totalsteps$V1,na.rm=TRUE)), col=c("red"),lwd=3,lty=3)

text(x=mean(totalsteps$V1,na.rm=TRUE),y=0,"10766",pos=3)

#Adding the legend for ease of interpretation
legend("topright","mean and median",col="red",lty=3)
`````` 
#### What is the average daily activity patterrn?

I will use the data.table package to get the average steps per interval.

```{r}
avgsteps <- data[,round(mean(steps,na.rm=TRUE)),interval]
head(avgsteps)
```

```{r} 
plot(x = avgsteps$interval,y=avgsteps$V1,type='l', xlab="Interval",ylab="Steps",main="Average Steps per 5 min interval")
```

```{r}
#subset the avgsteps dataset to get the maximum interval
max_interval <- avgsteps[which(avgsteps$V1==max(avgsteps$V1))] 
max_interval
```
Plotting the vertical line to show that the interval with the maximum number of steps is 835.
```{r}
#Subset the dataset to calculate the mean steps per interval
avgsteps <- data[,round(mean(steps,na.rm=TRUE)),interval]

plot(x = avgsteps$interval,y=avgsteps$V1,type='l', xlab="Interval",ylab="Steps",main="Average Steps per 5 min interval")

#Adding text to display the max interval
text(max_interval$interval,0,"835",pos=2)

#Adding the legend for ease of interpretation
legend("topright","Max Interval",col="red",lty=1)

#Adding the vertical line at the maximum interval 835
abline(v=max_interval$interval,col="red")
```

###   Imputing Missing Values
##### Total missing values in the dataset

```{r}
#Total missing values in the dataset
sum(is.na(data)) 
```


```{r}
#Total missing values in the interval column
(sum(is.na(data$interval))) 
```

```{r}
#Total missing values in the steps column
print(sum(is.na(data$steps))) 
```

To impute the values in the steps column I will fill in the missing values using the average steps for that interval.

I will use the avgsteps table as a lookup table. 
NA values in the 'steps' column of 'data' are replaced with the average steps for its corresponding interval.
```{r}
#Using dplyr, add the replace_steps column with filled in values
#If the value in the steps column is missing, use the match function to find its appropriate average steps
#Pass the new dataset to the mutate function using the pipe operator
#Drop the old steps column
no_na <- mutate(data,replace_steps = ifelse(is.na(steps),avgsteps$V1[match(interval,avgsteps$interval)],steps)) %>% mutate(steps=NULL)

#Compare the two datasets (imputed data(no_na) vs missing data(data))
print(head(no_na))
print(head(data))
```
#### Make a histogram of the total number of steps taken each day

```{r}
#Convert the no_na from a data.frame object to a data.table for ease of subsetting
no_na <- as.data.table(no_na) 

# Select all rows and sum the replace_steps column while grouping by date
plot_data <- no_na[,sum(replace_steps),date] 

#Create a histogram using the plot_data table
hist(plot_data$V1,xlab="Total Steps",col="lightblue", main="Distribution of Total Steps per Day") 

#Add a vertical line to indicte the mean
abline(v=c(mean(plot_data$V1,mean(totalsteps$V1,na.rm=TRUE))),col=c("blue","red"),lty=2)
text(x=mean(plot_data$V1),y=0,"10766",pos=3)

# Add the legend for ease of interpretation
legend("topright","Mean",col="blue",lty=2)

```

```{r}
print(mean(plot_data$V1))
print(median(plot_data$V1))
```

There is little to no difference between the means of the data with and without na values.
```{r}
#Imputed mean
mean(plot_data$V1) 

#Mean without na
mean(totalsteps$V1,na.rm=TRUE) 
```


#### Are there differences in activity patterns between weekdays and weekends?
```{r}
#Create a list of weekdays
days <- c("Monday","Tuesday","Wednesday","Thursday","Friday") 

#Check if each date is in the weekdays list and print weekday or weekends correspondingly
no_na <- mutate(no_na,type_of_day = ifelse(weekdays(no_na$date) %in% days,"weekday","weekend")) 

head(no_na)
```

```{r}
#Subset no_na by selecting the date and type_of day columns,
#Calculate the mean of the replace_steps column and group by interval
no_na <- as.data.table(no_na)
plotting_data <- no_na[,.(date,type_of_day,round(mean(replace_steps))),.(interval)]
head(plotting_data)
```
#### Finally, make a panel plot containing a time series plot of the 5-minute interval (x-axis) and the average number of steps taken
#### Average the steps across all weekday days or weekend days (y-axis).
```{r}
p <- plotting_data %>% ggplot(aes(interval,V3)) 
p + geom_line() + facet_grid(type_of_day~.,switch = "y") + ylab("Average Steps") + xlab("Interval") + ggtitle("Average Steps Taken per Day")
```

